---
title: "workshop"
author: "John Navarro"
date: "July 6, 2017"
output: pdf_document
---
Stochastic processes and stationarity
```{r}
Seeds<-(1:6)*10
Realizations<-rep(NA,6)

for (i in 1:6) {
  set.seed(Seeds[i])
  Realizations[i]<-rnorm(1)
}
rbind(Seeds,Realizations)
plot(Seeds,Realizations,pch=16,col=c("red","blue","green","purple","cyan","magenta"))
legend("bottomright",
       lty=1,
       legend=c("Outcome.1","Outcome.2","Outcome.3","Outcome.4","Outcome.5","Outcome.6"),
       lwd=2,
       col=c("red","blue","green","purple","cyan","magenta"))
```
Nowfor the same seeds generate 6different trajectories of stochastic process
```{r}
Realizations.process<-cbind(1:300,matrix(NA,300,6))
colnames(Realizations.process)<-c("Time",
                               "Trajectory.1",
                               "Trajectory.2",
                               "Trajectory.3",
                               "Trajectory.4",
                               "Trajectory.5",
                               "Trajectory.6")
Realizations.phases<-rep(NA,6)
for (i in 1:6) {
  set.seed(Seeds[i])
  Realizations.phases[i]<-rnorm(1)
  Realizations.process[,i+1]<-3+(Realizations.process[,1]+
                                   30*sin((Realizations.process[,1]+20*Realizations.phases[i]*pi)/20))
}
matplot(Realizations.process[,1],
        Realizations.process[,-1],
        type="l",
        col=c("red","blue","green","purple","cyan","magenta"),
        lty=1,
        xlab="Time",
        ylab="Trajectories",
        main="Trajectories for 6 Different Outcomes",
        lwd=2)
legend("topleft",
       lty=1,
       legend=colnames(Realizations.process)[-1],
       lwd=2,
       col=c("red","blue","green","purple","cyan","magenta"))
```

#### 2. ACF

Estimate ACF ofdecile10 ofCRSP portfolio from Jan1967 to Dec 2009 monthly.
Read the data from m-dec12910.txt
```{r}
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week3_TimeSeries"
da=read.table(paste(datapath,"m-dec12910.txt",sep="/"),header=T)
head(da)
dim(da)
# Select the Decile 10 returns
d10=da$dec10
dec10=ts(d10,frequency=12, start=c(1967,1))
# unneccesary    par(mfcol=c(1,1))
# Plot the time series and its ACF
plot(dec10, xlab="year", ylab="returns")
title(main="Simple returns")
acf(d10, lag=24)

```

Test H0: p12 = 0, againstHA:p12!=0 which represents an effect of persistent positive returns observed for stocks in January.
Conpare the tt statistic p1*2T^-1/2 with 5% critical value qnorm(0.975, 0.1)
```{r}
f1=acf(d10, lag=24)
f1$acf
tt=f1$acf[13]*sqrt(516)
c(tt, qnorm(0.975,0,1))

```
Since tt islarger than the critical value, then we should reject H0 and this means that ther is a seasonality at lag 12

Conduct BoxLjung test to check H0: all correlations =0
the data are returns of IBM stock from Jan 1967 to Dec 2009
Read the file, transform it into log returns, plot ACF and run BoxLjung test
```{r}
da=read.table(paste(datapath, "m-ibmsp6709.txt",sep="/"),header=T)
head(da)
dim(da)
# transform to log returns
ibm <- da$ibm
lnibm <- log(ibm+1)
# plot acf of returns and log returns....looks similar
plot(acf(ibm, lag=100))
acf(lnibm, lag=100)
# run Box Ljung test
Box.test(ibm, lag=12, type="Ljung")
Box.test(lnibm, lag=12, type="Ljung")


```

Since both pvalues returned are not significant, we cannot reject the H0 of no correlations in IBM stock returns

Check Box-Ljung on CRSP decile 10 portfolio
```{r}
Box.test(d10, lag=12, type="Ljung")
Box.test(d10, lag=24, type="Ljung")

```
Since both pvalues are small, we reject the H0, there exists serial correlation in the portfolio returns

### 3 AR Models

Consider the time series of GNP in file q-gnp4710.txt
```{r}
da=read.table(paste(datapath,"q-gnp4710.txt",sep="/"),header=T)
head(da)
# Create GNP and its log
G=da$VALUE
LG=log(G)
# Create the log returns
gnp=diff(LG)
# Create the time index
# Plot the growth and the log returns
dim(da)
# Create the time series
tdx=c(1:253)/4+1947
head(tdx)
# Plot original data  G vs tdx
plot(tdx, G, xlab="year", ylab="GNP", type="l")
# plot log diff vs tdx
plot(tdx[2:253], gnp, type="l", xlab="year",ylab="growth")
abline(h=0)
```
Almost all quarterly returns are positive
The largest negative return is in 2008, the Great Recession
Calculate ACF and PACF

```{r}
par(mfcol=c(1,2))
acf(gnp, lag=12)
pacf(gnp, lag=12)
```
Since we see exp decat in ACF, and PACF shows non zero coeff at lag 3, we will fit AR(3) model
```{r}
(m1 <- arima(gnp, order=c(3,0,0)))
```
The characteristic AR polynomial is
1 - 0.4386z - 0.2063z^2 + 0.156z^3 = 0

Check if the residuals behave like white noise

```{r}
tsdiag(m1,gof=12)
```
 since the pvalues are small for lags 9-12, means that with these number of lags selected, the hypothesis that all autocorrelations are equal to zero has to be rejected
 
 Create coefficients of the characteristic polynomial as:
 
```{r}
p1=c(1, -m1$coef[1:3])
```
Then find its roots
```{r}
(r1 <- polyroot(p1))
k=2*pi/acos(1.616116/1.832674) # compute length of the period
k
```
there are 3 roots of characteristic polynomial, one is real, the other 2 are complex. The real root, 

The US economy tends to have pseudo cycles with average period of 12.8 quarters
Check how th AR order would be selected automatically w AIC
```{r}
mm1 <- ar(gnp, method = "mle")
mm1$order
names(mm1)
print(mm1$aic, digits=3)
```
Plotting the AIC numbers
```{r}
aic=mm1$aic  # For plotting below.
length(aic)
plot(c(0:12),aic,type='h',xlab='order',ylab='aic')
lines(0:12,aic,lty=2)
```
### Model Checking
Read monthly value weighted return index fro CRSP in the file m-ibm3dx2608.txt
```{r}
vw=read.table(file=paste(datapath,"m-ibm3dx2608.txt",sep="/"),header=T)[,3]
head(vw)
```
Identify possible order
```{r}
par(mfrow=c(1,2))
acf(vw,lag=15)
pacf(vw,lag=15)
```
 Fit AR(3) model
 
```{r}
(m3 <- arima(vw,order=c(3,0,0)))
```
compute the intercept phi
```{r}
# intercept
(1-.1158+.0187+.1042)*mean(vw)
# standard error of residuals
sqrt(m3$sigma2)
# Use Box Ljung to see if residuals have autocorrelation
Box.test(m3$residuals,lag=12,type="Ljung")
```
The resulting p value is too large, so we cannot reject the hypothesis that all autocorrelations are equal to zero

Calculate adjusted p-valuefor the same statistic
```{r}
(pv=1-pchisq(16.35,9))
```
Model lag 2 may not be necessary, estimate model w only lags 1 and 3

```{r}
(m3 <- arima(vw, order=c(3,0,0),fixed=c(NA,0,NA,NA)))
(1-.1136+.1063)*.0089  # compute phi(0)
# compute resid stand error
sqrt(m3$sigma2)
#Ljung box test
Box.test(m3$residuals, lag=12,type = "Ljung")
#Adjust to 10
# Adjusted pvalue
pv <- 1-pchisq(16.83,10)
pv
```

#### 4. MA Models
Read the monthly returns of CFSP Equal Weighted Index
```{r}
da=read.table(file=paste(datapath,"m-ibm3dx2608.txt",sep="/"),header=T)
head(da)
ew=da$ewrtn
# Check ACF and PACF
par(mfrow=c(1,2))
acf(ew,lag=15)
pacf(ew,lag=15)

```
A candidate model is MA(9)
```{r}
m1 <- arima(ew, order=c(0,0,9))
m1
# put restrictions on the lags 2,4,5,6,7,8
m1 <- arima(ew, order=c(0,0,9), fixed = c(NA,0,NA,0,0,0,0,0,NA,NA))
m1
```

restrictions improveAIC measure of the model

```{r}
# Check standard deviation of the residuals
sqrt(m1$sigma2)
# Run box ljung test
Box.test(m1$residuals, lag=12, type="Ljung")
```
With unadjusted df we cannot reject the H0
```{r}
pv=1-pchisq(17.6,9)
pv
```
Now we see that the null hypothesis is rejected
use h=986 as origin for predictions to have an out of sample values to check the accuracy of predictions
```{r}
m1 <- arima(ew[1:986],order=c(0,0,9),fixed=c(NA,0,NA,0,0,0,0,0,NA,NA))
m1
```
Now predict the time series for 10 steps using the model m1
```{r}
prd <- predict(m1,10)
prd
```
```{r}
cbind(Actual=tail(ew,10),Predicted=as.vector(prd$pred),Standard.Error=as.vector(prd$se))
```
 

































```{r}

```

