---
title: "FinancialAnalysisHomework1Navarro"
author: "John Navarro"
date: "June 23, 2017"
output: pdf_document
---

### 1. Read Chapter 1 of the book
### 2. Download and analyze excess returns of S&P 500

Calculate continuous daily excess returns of SP500("^GSPC") for the period from 1/1/2014 until 12/31/2014 using overnight Fed Funds rates as risk-free rates.
```{r}
# Install packages
suppressWarnings(library(quantmod))
# Read in the data
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week1_Introduction"
FedFunds <- read.csv(file=paste(datapath, 'RIFSPFF.csv', sep='/'))
SP500 <- read.csv(file = paste(datapath, 'FSP500.csv', sep='/'))
tail(SP500)
# Calculate the simple daily return
SP500$simple.excess.return <- SP500$RIFSPFF_N.B - SP500$SP500Returns
#Calculate the continuous excess return
SP500.cont.excess.return <- diff(log(SP500$simple.excess.return))
# Plot the continuous excess return
plot(SP500.cont.excess.return, type="l", ylab="Excess returns of SP500", xlab = "Date", xaxt="n", main="Continuous Excess Daily Returns of S&P 500")
axis(1, at=10*(1:25), labels=SP500[10*(1:25),1])

```
## 3. Download and analyze exchange rates

Answer the smae questions as in Exercise 5 on page 37 as a refresher of statistical analysis skills. Try to do it without using R demo code from the book

1. Download GBP/USD and USD/JPY exchange rates from Oanda using quantmode
```{r}
getFX("GBP/USD",src="oanda", from= "2017-01-02", to= "2017-05-31")
getFX("USD/JPY")
```

