---
title: "Week 2 Assignment"
output: pdf_document
name: John Navarro
---
#John Navarro#

##Part 1. Generate uniformly distributed random numbers
#1.1 Use runif()

```{r}
##set specific seed, use runif() to create a sample of 0s and 1s with 1000
##datapoints
set.seed(15)
Sample<-runif(1000,0,1)
```

#1.2 Simulate Uniform Random Sample on [0,1] Using Random.org

```{r}
##install random package, set nFlips to 1000, use randomNumbers() function to
##create data set dataFromRandom, which is a series of 0s and 1s. print the 
##first 6 items in the dataset
library(random)
nFlips<-1000
dataFromRandom<-randomNumbers(n=nFlips, min=0, max=1, col=1, base=2, check=TRUE)
head(dataFromRandom)
```

Learn how to turn your sequence of {0,1} into uniform random numbers on [0,1]
```{r}
##install compositions package
suppressMessages(library(compositions))
##create a function called bin2dec that takes the input Bin.Seq, removes the 
##"," and uses function unbinary() to turn the string into a number
bin2dec<-function(Bin.Seq){
  unbinary(paste(Bin.Seq,collapse=""))  
}
bin2dec(c(1,1,1,1,1,0))
```

Turn the sequence of zeros and ones dataFromRandom of length 1000 into a matrix with 10 columns and 100 rows
```{r}
##Create a matrix of 10 col and 100 rows
Binary.matrix<-matrix(dataFromRandom,ncol=10)
head(Binary.matrix)
```

Transform each row of the matrix into decimal format using bin2dec() and divide the numbers by 2^10 to make real numbers in [0,1]
```{r}
##Use bin2dec() on each row and divide the values by 2^100 to get 100 decimal 
##values
Decimal.Sample<-apply(Binary.matrix,1,bin2dec)/2^10
Decimal.Sample
```

##Part 2. Test random number generator

#2.1 Test uniformity of distribution of both random number generators

2.1.1. Sample obtained by runif()

Analyze what was simulated by first looking at the histogram.
```{r}
##plot a histogram of Sample, the data from runif()
Sample.histogram<-hist(Sample)
Sample.histogram
```

*What does the histogram tell you about the distribution? Is it consistent with the goal of simulation?*
This histogram is a uniform distribution. Yes, it is consistent with the goal of the simulation. Which was to create a random uniform set of 1000 points all between 0 and 1.

Estimate mean and standard deviation of Sample.histogram$density.

```{r}
##find the mean and sd of the sample$density
(Sample.histogram.mean<-mean(Sample.histogram$density))
(Sample.histogram.sd<-sd(Sample.histogram$density))
##plot the histogram showing density not frequency
plot(Sample.histogram,freq=FALSE)
##draw the lines for the mean and the 95% Conf Intv
abline(h=Sample.histogram.mean)
abline(h=Sample.histogram.mean+1.96*Sample.histogram.sd,col="red",lty=2)
abline(h=Sample.histogram.mean-1.96*Sample.histogram.sd,col="red",lty=2)
```

*What does the graph tell you about the observed distribution?*

It looks like a uniform distribution with mean of 1 and a standard deviation of 0.0868

```{r}
##Find the mean and variance of the Sample
(Sample.mean<-mean(Sample))
(Sample.variance<-var(Sample))
```

*What do you conclude about the estimated distribution from the moments?*

The mean of Sample is 0.516 and the variance is 0.0841 so most of the data is centered around 0.5 with a narrow variance.

Check the summary of the simulated sample.
```{r}
#run the summary of the sample
summary(Sample)
```
*What do you think is the best way of estimating uniform distribution over unknown interval?*
The best way to estimate uniform distribution is by analyzing the moments.

2.1.2. Repeat the same steps to test uniformity of the sample created from Random.org data.
```{r}
##plot a histogram using the decimal values from bin2dec() function
D.Sample.histogram<-hist(Decimal.Sample)
D.Sample.histogram
#Calculate the mean and sd of histogram$density
(D.Sample.histogram.mean<-mean(D.Sample.histogram$density))
(D.Sample.histogram.sd<-sd(D.Sample.histogram$density))
##plot the histogram with density, not frequency
plot(D.Sample.histogram,freq=FALSE)
##plot the mean and Conf Intv
abline(h=D.Sample.histogram.mean)
abline(h=D.Sample.histogram.mean+1.96*D.Sample.histogram.sd,col="red",lty=2)
abline(h=D.Sample.histogram.mean-1.96*D.Sample.histogram.sd,col="red",lty=2)
##Calculate the mean, variance and summary of the sample
(D.Sample.mean<-mean(Decimal.Sample))
(D.Sample.variance<-var(Decimal.Sample))
summary(Decimal.Sample)
```
*What does the histogram tell you about the distribution? Is it consistent with the goal of simulation?*
That the distribution is not uniform, It looks as if the mode is around 0.2. It does not seem consistent with the goal of the simulation.
*What does the graph tell you about the observed distribution?*
It says that the density has a mean of 1, but a wide confidence interval, due to the large standard deviation.
*What do you conclude about the estimated distribution from the moments?*
This distribution has a similar mean, but a larger variance. It still gives a uniform distribution.
*What do you think is the best way of estimating uniform distribution over unknown interval?*

The best way to estimate uniform distribution is by analyzing the moments.

#2.2. Test independence of the sequence of zeros and ones
2.2.1. Turning point test
```{r}
##install randtests package, run turning.point.test() on decimal.sample
library(randtests)
turning.point.test(Decimal.Sample)
```

2.3. Test frequency by Monobit test

```{r}
##transorm data of {0,1} to {-1,1}
dataFromRandom.plusminus1<-(dataFromRandom-.5)*2
##create 2 complimentary functions erf and erfc, that 
erf <- function(x) 2 * pnorm(x * sqrt(2)) - 1
erfc <- function(x) 2 * pnorm(x * sqrt(2), lower = FALSE)
##plot the erfc
plot(seq(from=-3,to=3,by=.05),erfc(seq(from=-3,to=3,by=.05)),type="l",xlab="x",ylab="erfc(x")
```
To test the sequence RiRi check the value erfc(S).

If the P-value or erfc(S) is less or equal than 0.01 the sequence fails the test.
```{r}
##Sum the rows, and take absolute value and divide by 100 to get decimal
##Use erfc, to get the pvalues of the sub-sequences
##see which ones are less than 0.01 and sum the true ones
erfc(abs(sum(dataFromRandom.plusminus1)/sqrt(2*nFlips)))
```

The test shows that our sequence passes.

Now check each of the sub-sequences created earlier:
```{r}
plot(erfc(abs(apply(matrix(dataFromRandom.plusminus1,ncol=50),1,sum))/sqrt(2*50)),ylab="P-values of 20 runs")

```

*How many runs out of 20 fail the test?*

```{r}

sum(erfc(abs(apply(matrix(dataFromRandom.plusminus1,ncol=50),1,sum))/sqrt(2*50))<=.01)


```

None of them fail, there are none less than 0.01

##Part 3. Invent a random number generator
Think about possible sources of true or pseudo-random sequences of {0,1} and choose one or two of them.
Conduct the tests described in the previous section.

#3.1 Description of your random number generator
My sequence is taken from a MLB pitcher (Jake Arrieta) career statistics for walks and strikeouts per game. For each game I tabulated a ratio of walks divided by strikeouts, and removed the NA values(for when he had zero strikeouts) as well as the values greater than 1 for when he had more walks than strikeouts. We are left with 128 datapoints between 0 and 1.

#3.2 Generated sequence
```{r}
##read in the data
setwd("C:/Users/JohntheGreat/Documents/MSCA/StatisticalAnalysis/Week2/Assignments")
ratio <- read.csv(file = "ArrietaRatio.csv", header = F)
bbk.ratio <- ratio$V3
head(bbk.ratio)
```


#3.3 Uniformity test

```{r}
## plot a histogram of bbk,ratio
bbk.ratio.histogram<-hist(bbk.ratio)
bbk.ratio.histogram
##calc the mean and sd of the density
(bbk.ratio.histogram.mean<-mean(bbk.ratio.histogram$density))
(bbk.ratio.histogram.sd<-sd(bbk.ratio.histogram$density))
##plot the histogram and the mean and CI 
plot(bbk.ratio.histogram,freq=FALSE)
abline(h=bbk.ratio.histogram.mean)
abline(h=bbk.ratio.histogram.mean+1.96*bbk.ratio.histogram.sd,col="red",lty=2)
abline(h=bbk.ratio.histogram.mean-1.96*bbk.ratio.histogram.sd,col="red",lty=2)
##Calc the mean, variance and summary of bbk.ratio
(bbk.ratio.mean<-mean(bbk.ratio))
(bbk.ratio.variance<-var(bbk.ratio))
summary(bbk.ratio)
```
*What does the histogram tell you about the distribution? Is it consistent with the goal of simulation?*
The distribution is not uniform, it looks like the frequencies are high for low values, and low for the values between 0.5 and 1.

*What does the graph tell you about the observed distribution?*
It tells me that the variance of the densities are large, especially none of the data falls withing the confidence interval.

*What do you conclude about the estimated distribution from the moments?*
It tells me that that mean is at 0.42, and not close to 0.50. Most of the data is under 0.50

*What do you think is the best way of estimating uniform distribution over unknown interval?*
The best way to estimate uniform distribution is by analyzing the moments.

#3.4 frequency test

```{r}
library(compositions)
##use binary to convert bbk.ratio to binary values
bbk.bin <- binary(bbk.ratio * 100)
##paste all values to form a string of 0s and 1s
##split all the characters and turn them into numerics
bbk.sample <- as.numeric(unlist(strsplit(paste(bbk.bin, collapse = ""), split="")))
##turn the 0s and 1s to +/-1s
bbk.bin.plusminus1 <- (bbk.sample - .5)*2

erf <- function(x) 2 * pnorm(x * sqrt(2)) - 1
erfc <- function(x) 2 * pnorm(x * sqrt(2), lower = FALSE)

plot(seq(from=-3,to=3,by=.05),erfc(seq(from=-3,to=3,by=.05)),type="l",xlab="x",ylab="erfc(x")

erfc(abs(sum(bbk.bin.plusminus1)/sqrt(2*nFlips)))

```
Since the Pvalue is less than 0.01, by a signifcant amount, this sequence fails the test and is not random.
```{r}

##I only have 896 data points, so I used a matrix with 28 columns and 32 rows. 
plot(erfc(abs(apply(matrix(bbk.bin.plusminus1,ncol=28),1,sum))/sqrt(2*250)),ylab="P-values of 32 runs")
##Sum the rows, and take absolute value and divide by 100 to get decimal
##Use erfc, to get the pvalues of the sub-sequences
##see which ones are less than 0.01 and sum the true ones
sum(erfc(abs(apply(matrix(bbk.bin.plusminus1,ncol=28),1,sum))/sqrt(2*50))<=.01)
```
None of the sub sequences are less than 0.01

#3.5 Turning point test

```{r}
##install randtests package, run turning.point.test() on decimal.sample
library(randtests)
turning.point.test(bbk.ratio)
```

##Part 4. Monte Carlo Method

#4.1. Scratch off quote of the day: function download
```{r}
##load the file
datapath<-"C:/Users/JohntheGreat/Downloads/documents%2FMScA Statistical Analysis 31007%2FMScA 31007 Lecture 2%2FScratchOffMonteCarlo (1).rda/"
#load(file=paste(datapath,'ScratchOffMonteCarlo.rda',sep='/'))

##load(file = datapath)
load("C:/Users/JohntheGreat/Downloads/documents%2FMScA Statistical Analysis 31007%2FMScA 31007 Lecture 2%2FScratchOffMonteCarlo.rda")
```

#4.2. Simulate pseudo-random poins [x,y][x,y] on [0,100]×[0,100]

Select a number o points nSample.

Simulate a sample of length 2*nSample from uniform distribution on [0,100] and turn it into a (nSample×2)(nSample×2) matrix.
Use a seed of your choice my.seed.
```{r}
##Set sample and seed, use runif to create random values from 0 to 100 and put 
##into a matrix with two columns
nSample <- 20000
my.seed <- 5555
set.seed(my.seed)
xy<-runif(2*nSample,0,100)
xy<-matrix(xy,ncol=2)
head(xy)
```

Throw nSample simulated points on square [0,100]×[0,100][0,100]×[0,100] to scratch off some of yellow paint.
```{r}
##run scratchoff function using the matrix xy
ScratchOffMonteCarlo(xy)
```
Take a note of the percentage scratched off returned by ScratchOffMonteCarlo(xy)

*By changing nSample and my.seed try to make the quote of the day readable with minimum sample size.
What percent you needed to scratch off to make the quote readable?*

I needed over 86% to make the quote readable, which was nSample <- 20000

##4.3. Simulate quasi-random poins [x,y][x,y] on [0,100]×[0,100]

```{r}
##install randtoolbox package
library(randtoolbox)
```
Run sobol() first time with the default set for parameter init=T.
```{r}
##set seed and nsample
my.seed<-6565
set.seed(my.seed)
nSample<-10
## use sobol function to create a 2 column matrix with 10 rows
xy<-sobol(nSample,dim=2,init=T)*100
```

Then make init=F if you want to generate different sample every time or keep it equal to T if you want repeated samples.

```{r}
##run using a different seed and sample
my.seed<-123
nSample<-13000
xy<-sobol(nSample,dim=2,init=T,scrambling = T,seed=my.seed)*100
##Plot the data pairs and run ScratchOffMonteCarlo
plot(xy)
ScratchOffMonteCarlo(xy)
```

*Again, by changing nSample and my.seed try to make the quote of the day readable with minimum sample size.*
*What percent you needed to scratch off to make the quote readable?*
Again, I needed about 86% open to make it readable

*Which of the Monte Carlo methods makes the quote readable sooner?*
The Sobol method was able to reach a readable percentage with a smaller nSample size

*Which parameters nSample and my.seed gave you the best result, what percent of the yellow paing you were able to scratch off by each method?*
For runif(), I used my.seed <- 5555 with nSample <- 20000
For Sobol, I used my.seed <- 123 with nSample <- 13000
Both got the percent open to around 86%
*Changing which of the two parameters plays more significant role?*
Changing nSample played a more significant role.


