---
title: "Week2AssignmentPart3"
output: pdf_document
---

##Part 3. Invent a random number generator
Think about possible sources of true or pseudo-random sequences of {0,1} and choose one or two of them.
Conduct the tests described in the previous section.

#Part1 Description of your random number generator
My sequence is taken from a MLB pitcher (Jake Arrieta) career statistics for walks and strikeouts per game. For each game I tabulated a ratio of walks divided by strikeouts, and removed the NA values(for when he had zero strikeouts) as well as the values greater than 1 for when he had more walks than strikeouts. We are left with 128 datapoints between 0 and 1.

#Part2 Generated sequence
```{r}
setwd("C:/Users/JohntheGreat/Documents/MSCA/StatisticalAnalysis/Week2/Assignments")
ratio <- read.csv(file = "ArrietaRatio.csv", header = F)
bbk.ratio <- ratio$V3
head(bbk.ratio)
```


#Part3 Uniformity test

```{r}
bbk.ratio.histogram<-hist(bbk.ratio)
bbk.ratio.histogram
(bbk.ratio.histogram.mean<-mean(bbk.ratio.histogram$density))
(bbk.ratio.histogram.sd<-sd(bbk.ratio.histogram$density))
plot(bbk.ratio.histogram,freq=FALSE)
abline(h=bbk.ratio.histogram.mean)
abline(h=bbk.ratio.histogram.mean+1.96*bbk.ratio.histogram.sd,col="red",lty=2)
abline(h=bbk.ratio.histogram.mean-1.96*bbk.ratio.histogram.sd,col="red",lty=2)
(bbk.ratio.mean<-mean(bbk.ratio))
(bbk.ratio.variance<-var(bbk.ratio))
summary(bbk.ratio)
```

#Part4 frequency test

```{r}
##convert bbk.ratio to binary sample bbk.bin
##For now, use runif for set of {0,1}
nFlips <- 1000
set.seed(15)
bbk.bin<-runif(1000,0,1)

bbk.bin.plusminus1 <- (bbk.bin - .5)*2

erf <- function(x) 2 * pnorm(x * sqrt(2)) - 1
erfc <- function(x) 2 * pnorm(x * sqrt(2), lower = FALSE)

plot(seq(from=-3,to=3,by=.05),erfc(seq(from=-3,to=3,by=.05)),type="l",xlab="x",ylab="erfc(x")

erfc(abs(sum(bbk.bin.plusminus1)/sqrt(2*nFlips)))

plot(erfc(abs(apply(matrix(bbk.bin.plusminus1,ncol=50),1,sum))/sqrt(2*50)),ylab="P-values of 20 runs")

sum(erfc(abs(apply(matrix(bbk.bin.plusminus1,ncol=50),1,sum))/sqrt(2*50))<=.01)
```

#Part5 Turning point test

```{r}
library(randtests)
turning.point.test(bbk.ratio)
```

