---
title: "FA Case Study Week 4"
author: "John Navarro"
date: "July 13, 2017"
output: pdf_document
---

# Retail Gas Case Study
## 1.1 Data
```{r}
suppressWarnings(library(xts))
suppressWarnings(library(quantmod))
gasoline<-read.csv(file=paste(datapath,"Weekly U.S. Regular All Formulations Retail Gasoline Prices.csv",sep="/"))
head(gasoline)
```

transorm into time series
```{r}
g<-xts(gasoline$Gasoline,order.by=as.Date(gasoline$Date,format="%m/%d/%y"))
plot(log(g))
dg<-diff(log(g))[-1]
plot(dg)
```
 
 ## 1.2 Tasks
 1. Analyze ACF and PACF, list candidates
 
```{r}
par(mfrow=c(1,2))
acf(dg, lag=50)
pacf(dg, lag = 50)
```
AR model 3, bc of PACF
2. Use function ar() to identify AR order

```{r}
ar(dg) # suggests AR(23)
auto.arima(dg, max.p = 10,max.q = 10, max.order = 30) # suggests(2,0,3)
eacf(dg)
```
3. Fit all candidates, look at AIC
```{r}
(m3 <- arima(dg, order=c(3,0,0))) #AIC -7273.94
(m23<- arima(dg, order=c(23,0,0)))# AIC -7286.77
(m203<- arima(dg, order=c(2,0,3))) # AIC -7281.15
```
4. Analyze signif of estimates of coeff and identify a reduced model w fixed insig coef.
Fit this model
```{r}
m23.b <- arima(dg, order=c(23,0,0), fixed = c(NA,NA,NA,0,NA,0,0,NA,NA,NA,0,0,0,NA,NA,0,NA,NA,NA,0,0,0,NA,NA), transform.pars = F)
m23.b  # AIC -7299.04

```
5. Run tsdaig() for all candidates, compare the candidates by number of lags for which B_L NH is rejected
```{r}
tsdiag(m3)
tsdiag(m23)
tsdiag(m203)
tsdiag(m23.b)
```
6. Compare the candidates by pvalues of Box-Ljung test using Box.test()
```{r}
Box.test(m3$residuals, type='Ljung') #pv = 0.875
Box.test(m23$residuals, type='Ljung') #pv = 0.9524
Box.test(m203$residuals, type='Ljung') #pv = 0.9959
Box.test(m23.b$residuals, type='Ljung') #pv = 0.8728
```
7. Manually adjust p-values of BL test for residual # dgress of freedom
```{r}


```
8. Select the best candidate explain


m23, with the fixed insig. coeff has the lowest AIC....BOOM!

## 1.3 Use the fitted model for prediction of log retursn for 10 weeks starting from 2/13/12

### 1.3.1 Create short time series that ends on 2/13/12 - forecast origin
```{r}
dgShort <- diff(log(first(g,n=1100)))[-1]
plot(dgShort)
head(dgShort)
tail(dgShort)

```
### 1.3.2 Fit the best candidate model to the short time series to create predictiveModel


