---
title: "Workshop2"
author: "John Navarro"
date: "July 13, 2017"
output: pdf_document
---

# Intervention Model


```{r, warning=FALSE, include=FALSE}
library(tseries)
library(forecast)
library(TSA)
```


use on pulse as input into transition function.
What comes out, something trivial
shock, then exponential calming down

```{r}
dat <- read.table(paste(datapath,'q-ko-earns8309.txt',sep = '/'), header=TRUE)
head(dat)
# Make time series of earnings
par(mfrow = c(1, 1), cex = 1.0)
ko <- ts(dat$value,start=1983,frequency =4)
plot(ko,type = "b",pch = 21,lwd = 2, col="blue")
time(ko)

# Take logs
logko <- log(ko)
plot(logko, col="blue", type="b", lwd=2)
```

## 2.2 Observation of intervention

Observe change in the trend on late 1990s
The suspected intervention moment is last quarter of 1998
```{r}
MT <- 63
(Intervention.Time <- 1983+(MT-1)*0.25)
ko
```
Plot the logged data with suspected intervention quarter marketd
```{r}
plot(logko, col = "blue",type = "b",lwd =2)
abline(v = Intervention.Time,col = "red",lwd = 1)
text(x = Intervention.Time-2,y = -2.8,labels = as.character(Intervention.Time-.25),col = "red")
```
To elminate trend take simple differences and seasonal differences of period 4
```{r}
dko <- diff(logko) # first difference
ddko <- diff(dko,4) # seasonal difference
plot(dko, col="blue", type="b", lwd=2)
abline(v=Intervention.Time, col="red", lwd=1)
text(x = Intervention.Time-2, y= -0.3, labels= (Intervention.Time),col="red")
```

After taking simple differences there is still significant seasonal component in the data
```{r}
plot(ddko, col="blue", type="b",lwd=2)
abline(v=Intervention.Time, col="red")
text(x = Intervention.Time-2, y=-0.3, labels=Intervention.Time, col="red")
```

## 2.3 SARMA model Identification

Observe ACF and PACF of simple and seasonally differenced data
```{r}
acf(ddko, lag.max = 48, lwd=4, col="blue", main="ACF")
pacf(ddko, lag.max=20, lwd=4, col="blue", main="PACF")
```
### 2.3.1 SARIMA Model

## 2.4 SARIMA and Intervention model estimation

```{r}
iAR <- 1
(I1 <- 1*(seq(ko)>=MT))
ko.sarima=arimax(logko, order=c(1,1,0),
                 seasonal = list(order=c(2,1,0),period=4),
                 xtransf=data.frame(I1,I1),
                 transfer=list(c(0,0),c(iAR,0)), method = "ML")
#####################what is transfer=################################

AIC(ko.sarima)
  
```

## 2.5 Fitting results

Model estimates
```{r}
ko.sarima$coef
(u_0 <- ko.sarima$coef[4])
(w_0 <- ko.sarima$coef[(5+iAR)])
(r_1 <- ko.sarima$coef[(5+iAR-1)])
plot(logko,type = "p", pch = 20,ylab='Log(KO)',main = 'Fitting results',col= "blue",lwd =2)
points(fitted(ko.sarima),col="orange",lwd=2)
```

Residuals

```{r}
plot(ko.sarima$residuals, type='b',pch=20,main='Residuals',col='blue',lwd=2)
abline(v=Intervention.Time, col='red',lwd=1)
text(x=Intervention.Time,y=-0.18,labels=as.character(Intervention.Time),col="red")

```

## 2.6 Show intervention effect

Plot the estimated component mt

```{r}
#################### what is this equation?###################333
inter1 <- I1*u_0+filter(I1,filter=r_1,method='recursive',side=1)*(w_0)
inter1 <- ts(inter1,start=1983, frequency = 4)
plot(inter1,ylab='Intervention Effects',type='l',col='blue',lwd=2)
abline(h=0)
```

## 2.7 Remove intervention effect from the data

```{r}
logko.interven.deleted <- logko - inter1
plot(logko.interven.deleted,
     main= paste('Log(KO).Intervention',
                 Intervention.Time, 'deleted'),
      col= 'blue', type='b', lwd=2)
abline(v=Intervention.Time, col='orange',lwd=1)
text(x=Intervention.Time-2, y=-3.0, labels= Intervention.Time, col="red")
```

## 2.8 Discover another intervention

The plot of the data after the estimated intervention has been removed shows another shock that finally ends the episode: aftershock or correction

```{r}
plot(logko.interven.deleted,
     main=paste('Log(KO).Intervention',
                Intervention.Time,'deleted'),
     col='blue',type='b',lwd=2)
abline(v=Intervention.Time,col='orange', lwd=1)
text(x=Intervention.Time-2,y=-3.0,labels= Intervention.Time, col='red')
abline(v=Intervention.Time+1.5, col="green",lwd=1)
text(x=Intervention.Time+3,y=-3.0,labels=Intervention.Time+1.5,col="green")
```

All equations for the second intervention are the same. But as the second intv, we fit pulse function
```{r}
MT2 <- 69
(Intervention.Time2 <- 1983+(MT2-1)*0.25)
iAR <- 1
(I2 <- 1*(seq(ko)==MT2))
### How does this work?#####
```

```{r}
ko.sarima2=arimax(logko.interven.deleted,
                   order=c(1,1,0),
                   seasonal=list(order=c(2,1,0),period=4),
                   xtransf=data.frame(I2,I2),
                   transfer=list(c(0,0),c(iAR,0)),method='ML')
ko.sarima2$coef
```



## 2.9 Calculate Intervention effect
```{r}
(u_0 <- ko.sarima2$coef[4])
(r_1 <- ko.sarima2$coef[5:(5+iAR-1)])
(w_0<-ko.sarima2$coef[(5+iAR)])
p2=1*(seq(ko)==MT2)
inter2<- p2*u_0+ filter(p2,filter=r_1,method='recursive', side=1)*
  (w_0)
inter2 <- ts(inter2,start = 1983,frequency = 4)

plot(inter2,main = paste("Intervention Effect. Intervention Time=",Intervention.Time2), ylab='Intervention Effects',
     type='l',col = "blue",lwd = 2)
abline(h=0)
```

## 2.10 REmove second intervention

```{r}
logko.interven.deleted.inter2 <- logko.interven.deleted - inter2
plot(logko.interven.deleted.inter2,main = paste("Logko. Interventions ",Intervention.Time,"and",Intervention.Time2, " deleted"), col = "blue",type = "b",lwd =2)
abline(v= Intervention.Time,col = "red",lwd = 1)
abline(v= Intervention.Time2,col = "green",lwd = 1)
text(x = Intervention.Time-2,y = -3,labels = as.character(Intervention.Time), col = "red")
text(x = Intervention.Time2+2,y = -2,labels = as.character(Intervention.Time2), col = "green")
```





