---
title: "Navarro_Assignment3_b"
author: "John Navarro"
date: "January 28, 2017"
output: pdf_document
---

# 1. Split the sample into two random samples. Standardize the data
```{r}
#Load the German data
#germanCredit<- read.csv("C:/Users/JohntheGreat/Documents/MSCA/DataMining/Week1/german_credit.csv", header=TRUE, sep=",")
library(caret)
data(GermanCredit)
head(GermanCredit)
# select only the chosen variables
GCD<- GermanCredit[, c(1:7)]
#colnames(GCD)
#summary(GCD)
# need to split the data into train/test portions
set.seed(555)
train_ind <- sample(seq_len(nrow(GCD)), size = 700)
# separate into two data frames: train and test
train_data <- GCD[train_ind, ]
test_data<- GCD[-train_ind, ]
head(train_data)
head(test_data)
#standardize train_data
train_means = colMeans(train_data)
train_sd = apply(train_data, 2, sd)
train_scaled = scale(train_data)
head(train_scaled)
test_scaled = scale(test_data, center=train_means, scale = train_sd)
head(test_scaled)
```

# 2. Perform principal components of variables 1:7 on the training sample

```{r}
pr.cr <- prcomp(train_scaled)
summary(pr.cr)
attributes(pr.cr)
pr.cr
#Loadings/coefficients
round(pr.cr$rotation,2)
#components
round(head(pr.cr$x),2)

```

# 3. Generate Scree Plots and select the number of components to retain
```{r}
plot(pr.cr$sdev^2, ylab= "Variances", type='b')
#Variance Explained
pr.var <- pr.cr$sdev ^2
pve <- pr.var/sum(pr.var)
plot(pve, xlab="Principal Component", ylab="Proportion of Variance Explained ", ylim=c(0,0.3), type = "b")
plot(cumsum(pve), xlab="Principal Component ", ylab=" Cumulative Proportion of Variance Explained ", ylim=c(0,1), type = "b")
```
It appears that there is a bend at the 3rd Principal Component.

# 4. Plot Component 1 loadings vs Component 2 loadings

```{r}
plot(x=pr.cr$rotation[, 1], y=pr.cr$rotation[, 2], xlab='PC1 Loadings', ylab='PC2 Loadings', main = 'PC1 v PC2')
plot(x=pr.cr$rotation[, 1], y=pr.cr$rotation[, 3], xlab='PC1 Loadings', ylab='PC3 Loadings', main = 'PC1 v PC3')
#Loadings/coefficients
round(pr.cr$rotation[,1:3],2)
```

From the PC plots and from the PC loadings we can see how they are related. PC1 seems to have a strong negative relationship with Loan Duration and Credit Amount. Therefore, we can describe PC1 by the "Total Credit"" (a representation of the amount combined with the duration). PC2 has the strongest relationship with (in descending order) Age, Residence duration, Number of exisisting credits at the bank, and Number of people being liable to provide maintenance for. These attributes could be categorized under the name "Developed Household". For instance, the longer you have lived at a residence, the older you usually are, the more loans you have taken out, and the more people are in your family. PC3 Is most strongly influenced by Installment Rate in percentage of disposable income, so we can call it "Loan Rate."


# 5. Show that the Loadings are orthogonal


