---
title: "Currency_project"
author: "John Navarro"
date: "February 11, 2017"
output: pdf_document
---

# Objective

The objective of this research is to classify daily FX data into currency groupings and create trading strategies within these groupings. 

# Collecting the Data

We have chosen to analyze the top 13 traded FX currencies along with spot gold. The dataset is composed of 3 years of daily closing prices. By looking at the data, we see that the currency values are very different. For instance, the South Korean won closed at 1.4327 while the US dollar trades at 100.80. In order to compare daily changes across countries, we need to normalize the data. We chose to represent the daily values as a percent change. 
(Today's price - Yesterday's price) / Yesterday's price

```{r}
# Import the data
# This is the closing price data
CurrClose<- read.csv("C:/Users/JohntheGreat/Documents/MSCA/DataMining/CourseProject/Currencies_close_data.csv", header=TRUE, sep=",")

# This is the data after changing to percent change
CurrDataPer<- read.csv("C:/Users/JohntheGreat/Documents/MSCA/DataMining/CourseProject/Currencies_percent_change.csv", header=TRUE, sep=",")

# A quick look at the standardized data
summary(CurrDataPer)
head(CurrDataPer)

# The data must be transposed to use hclust()
CurrData<- read.csv("C:/Users/JohntheGreat/Documents/MSCA/DataMining/CourseProject/Currencies_percent_change_transposed.csv", header=TRUE, sep=",")
```

Separate all three data sets into Training and Test sets. Since this is time series data, we will keep the data sets continuous. The Training set will contain the first 68% of the data while the test set will include the final 32% of the data.

```{r}
# This is the transposed data for the clustering calculations
CurrData.Train <- CurrData[,2:501]
CurrData.Test <- CurrData[ ,502:734 ]

#This is the closing price data to be used in the regression calculations
CurrDataPer.Train <- CurrDataPer[1:500,]
CurrDataPer.Test <- CurrDataPer[501:733,]

# This is the actual close prices to be used with the trading signals
CurrClose.Train <- CurrClose[1:500,]
CurrClose.Test <- CurrClose[501:733,]


```

#Using Hierarchical Clustering to find similar pairs

```{r}
# Separate by clusters using 4 different methods
clusters <- hclust(dist(CurrData.Train), method = 'ward.D')
plot(clusters, main= "Ward dendogram Training")

clusters <- hclust(dist(CurrData.Train), method = 'single')
plot(clusters, main= "Single Link dendogram Training")

clusters <- hclust(dist(CurrData.Train), method = 'complete')
plot(clusters, main= "Complete Link dendogram Training")

clusters <- hclust(dist(CurrData.Train), method = 'average')
plot(clusters, main= "Average Link dendogram Training")
```


From the dendograms of the full set of currencies, we see the following groupings are consistently paired:

+ Commodity currencies-(4-AUD and 6-NZD)
+ European currencies-(1-EUR and 9-EUR/GBP)
+ Asian Currencies-(13-CNY and 14-HKD).
+ North American Currencies(5-CAD and 8-MXN)
+ Yen and US(2-JPY and 11-USD)

Four of these groupings make sense anecdotally. Australia and New Zealand's economies and currencies are closely tied together and were typically involved in the carry trade. Similarly the economies of China and Hong Kong are very correlated, we can see this in their currency relationship. Canada and Mexico are 2 of the top 3 biggest trading partners with the US. Japan and the US have had low interest rates during this time period. They both have been used as funding currencies for the carry trade. 

9-EUR/GBP is widely traded, but is obviously associated with one of its components 1-EUR. 10-GOLD also seems to be the first product to be split off, it is consisitently dissimilar from the other products.  We can remove these two from the data set and see if the groupings change. Unfortunately, we cannot pair anything with the US dollar. Since currencies have to be traded in pairs, most settle in dollars. We will also remove 11-USD.


```{r}
# Remove 8,9,and 10 from the clusters
CurrData2.Train <- CurrData.Train[c(1:8,12:14), ]

# Separate by clusters using different methods
clusters <- hclust(dist(CurrData2.Train), method = 'ward.D')
plot(clusters, main= "Ward dendogram Training")

clusters <- hclust(dist(CurrData2.Train), method = 'single')
plot(clusters, main= "Single Link dendogram Training")

clusters <- hclust(dist(CurrData2.Train), method = 'complete')
plot(clusters, main= "Complete Link dendogram Training")

clusters <- hclust(dist(CurrData2.Train), method = 'average')
plot(clusters, main= "Average Link dendogram Training")
```

Here we see that the pairs (5,8),(4,6) and (13,14) are consistently similar in the various dendogram methods. Pair (1,3) appears in the ward dendogram which consists of the European Currencies (1-EUR and 3- GBP).

Moving forward, we can look at regression and correlation between these specific pairs to create a simple trading strategy.

We could also, cut the dendogram higher and look at larger groupings. For instance in the Ward Dendogram, we can see that(1,3,4,6) are in one group. While (7,5,8,12,13,14,2,11) appear to be in a larger group.

Alternatively, we could also try to model these as groups, in addition to pairs.

# Regression and Pairs Trading

```{r}
# Plots of the four pairs

plot(CurrDataPer.Train$EUR, CurrDataPer.Train$GBP, main= "Pound v Euro Training", xlab = "Euro", ylab="Pound")

plot(CurrDataPer.Train$NZD, CurrDataPer.Train$AUD, main= "Australia v New Zealand Training", xlab = "New Zealand", ylab="Australia")

plot(CurrDataPer.Train$HKD, CurrDataPer.Train$YUAN, main= "Renminbi v HK Dollar Training", xlab = "Hong Kong", ylab="Renminbi")

plot(CurrDataPer.Train$CAD, CurrDataPer.Train$MXN, main= "Mexico v Canada Training", xlab = "Canada", ylab="Mexico")

```



# Linear Regression of British Pound v Euro Training Set

```{r}

GBP.EUR <- lm(CurrDataPer.Train$GBP ~ CurrDataPer.Train$EUR-1)
summary(GBP.EUR)
# Look at the residuals
GBP.EUR.res <- GBP.EUR$residuals
hist(GBP.EUR.res,20)
par(mfrow=c(1,1))
plot(GBP.EUR.res, col = "blue", main = "Residual Plot Pound vs Euro Training")
acf(GBP.EUR.res)

# extract the hedge ratio
GBP.EUR.hr <- as.numeric(GBP.EUR$coefficients[1])

# Create the spread(percent change data)
GBP.EUR.spread.train <- CurrDataPer.Train$GBP - (GBP.EUR.hr * CurrDataPer.Train$EUR)

# compute the statistics of the spread (percent change data)
GBP.EUR.mean.train <- as.numeric(mean(GBP.EUR.spread.train, na.rm=T))
GBP.EUR.sd.train <- as.numeric(sd(GBP.EUR.spread.train, na.rm = T))
GBP.EUR.upperThr <- GBP.EUR.mean.train + 1.96 * GBP.EUR.sd.train
GBP.EUR.lowerThr <- GBP.EUR.mean.train - 1.96 * GBP.EUR.sd.train

# Visualize the spread (percent change) and thresholds
plot(GBP.EUR.spread.train, main = "Pound vs Euro spread (per change) training set", type = "l")
abline(h=GBP.EUR.mean.train, col = "red", lwd = 2)
abline(h=GBP.EUR.upperThr, col="blue", lwd=2)
abline(h=GBP.EUR.lowerThr, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high <- GBP.EUR.spread.train[GBP.EUR.spread.train > GBP.EUR.upperThr]
low <- GBP.EUR.spread.train[GBP.EUR.spread.train < GBP.EUR.lowerThr]
GBP.EUR.buys <- which(GBP.EUR.spread.train %in% high)
GBP.EUR.sells <- which(GBP.EUR.spread.train %in% low)

# Create the spread in price terms
GBP.EUR.spread.price.train <- CurrClose.Train$GBP - GBP.EUR.hr * CurrClose.Train$EUR

# Plot the spread(price) chart with buys and sells marked
plot(GBP.EUR.spread.price.train, main = "Pound v Euro spread (price) training set", type = "l")
abline(v=GBP.EUR.sells, col='red')
abline(v=GBP.EUR.buys, col='green')

```


# Linear Regression of Australian Dollar and New Zealand Dollar

```{r}

AUD.NZD <- lm(CurrDataPer.Train$AUD ~ CurrDataPer.Train$NZD-1)
summary(AUD.NZD)
# Look at the residuals
AUD.NZD.res <- AUD.NZD$residuals
hist(AUD.NZD.res,20)
par(mfrow=c(1,1))
plot(AUD.NZD.res, col = "blue", main = "Residual Plot Austrailia vs New Zealand Training")
acf(AUD.NZD.res)

# extract the hedge ratio
AUD.NZD.hr <- as.numeric(AUD.NZD$coefficients[1])

# Create the spread(percent change data)
AUD.NZD.spread.train <- CurrDataPer.Train$AUD - (AUD.NZD.hr * CurrDataPer.Train$NZD)

# compute the statistics of the spread (percent change data)
AUD.NZD.mean.train <- as.numeric(mean(AUD.NZD.spread.train, na.rm=T))
AUD.NZD.sd.train <- as.numeric(sd(AUD.NZD.spread.train, na.rm = T))
AUD.NZD.upperThr <- AUD.NZD.mean.train + 1.96 * AUD.NZD.sd.train
AUD.NZD.lowerThr <- AUD.NZD.mean.train - 1.96 * AUD.NZD.sd.train

# Visualize the spread (percent change) and thresholds
plot(AUD.NZD.spread.train, main = "Australia v New Zealand spread (per change) training set", type = "l")
abline(h=AUD.NZD.mean.train, col = "red", lwd = 2)
abline(h=AUD.NZD.upperThr, col="blue", lwd=2)
abline(h=AUD.NZD.lowerThr, col="blue", lwd=2)

# Pull the dates of the highest and lowest levels of spread (percent change)
high <- AUD.NZD.spread.train[AUD.NZD.spread.train > AUD.NZD.upperThr]
low <- AUD.NZD.spread.train[AUD.NZD.spread.train < AUD.NZD.lowerThr]
AUD.NZD.buys <- which(AUD.NZD.spread.train %in% high)
AUD.NZD.sells <- which(AUD.NZD.spread.train %in% low)

# Create the spread in price terms
AUD.NZD.spread.price.train <- CurrClose.Train$AUD - AUD.NZD.hr * CurrClose.Train$NZD

# Plot the spread(price) chart with buys and sells marked
plot(AUD.NZD.spread.price.train, main = "Australia v New Zealand spread (price) training set", type = "l")
abline(v=AUD.NZD.sells, col='red')
abline(v=AUD.NZD.buys, col='green')

```

# Linear Regression of Chinese Renminbi and Hong Kong Dollar

```{r}


CNY.HKD <- lm(CurrDataPer.Train$YUAN ~ CurrDataPer.Train$HKD-1)
summary(CNY.HKD)
# Look at the residuals
CNY.HKD.res <- CNY.HKD$residuals
hist(CNY.HKD.res,20)
par(mfrow=c(1,1))
plot(CNY.HKD.res, col = "blue", main = "Residual Plot Renminbi vs Hong Kong Dollar Training")
acf(CNY.HKD.res)

# extract the hedge ratio
CNY.HKD.hr <- as.numeric(CNY.HKD$coefficients[1])

# Create the spread(percent change data)
CNY.HKD.spread.train <- CurrDataPer.Train$YUAN - (CNY.HKD.hr * CurrDataPer.Train$HKD)

# compute the statistics of the spread (percent change data)
CNY.HKD.mean.train <- as.numeric(mean(CNY.HKD.spread.train, na.rm=T))
CNY.HKD.sd.train <- as.numeric(sd(CNY.HKD.spread.train, na.rm = T))
CNY.HKD.upperThr <- CNY.HKD.mean.train + 1.96 * CNY.HKD.sd.train
CNY.HKD.lowerThr <- CNY.HKD.mean.train - 1.96 * CNY.HKD.sd.train

# Visualize the spread (percent change) and thresholds
plot(CNY.HKD.spread.train, main = "Renminbi vs Hong Kong Dollar spread (per change) training set", type = "l")
abline(h=CNY.HKD.mean.train, col = "red", lwd = 2)
abline(h=CNY.HKD.upperThr, col="blue", lwd=2)
abline(h=CNY.HKD.lowerThr, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high <- CNY.HKD.spread.train[CNY.HKD.spread.train > CNY.HKD.upperThr]
low <- CNY.HKD.spread.train[CNY.HKD.spread.train < CNY.HKD.lowerThr]
CNY.HKD.buys <- which(CNY.HKD.spread.train %in% high)
CNY.HKD.sells <- which(CNY.HKD.spread.train %in% low)

# Create the spread in price terms
CNY.HKD.spread.price.train <- CurrClose.Train$YUAN - CNY.HKD.hr * CurrClose.Train$HKD
# Plot the spread(price) chart with buys and sells marked
plot(CNY.HKD.spread.price.train, main = "Renminbi vs Hong Kong Dollar spread (price) training set", type = "l")
abline(v=CNY.HKD.sells, col='red')
abline(v=CNY.HKD.buys, col='green')

```

#Linear Regression of Mexican Peso v Canadian Dollar

```{r}

MXN.CAD <- lm(CurrDataPer.Train$MXN ~ CurrDataPer.Train$CAD-1)
summary(MXN.CAD)
# Look at the residuals
MXN.CAD.res <- MXN.CAD$residuals
hist(MXN.CAD.res,20)
par(mfrow=c(1,1))
plot(MXN.CAD.res, col = "blue", main = "Residual Plot Mexican Peso vs Canadian Dollar Training")
acf(MXN.CAD.res)

# extract the hedge ratio
MXN.CAD.hr <- as.numeric(MXN.CAD$coefficients[1])

# Create the spread(percent change data)
MXN.CAD.spread.train <- CurrDataPer.Train$MXN - (MXN.CAD.hr * CurrDataPer.Train$CAD)

# compute the statistics of the spread (percent change data)
MXN.CAD.mean.train <- as.numeric(mean(MXN.CAD.spread.train, na.rm=T))
MXN.CAD.sd.train <- as.numeric(sd(MXN.CAD.spread.train, na.rm = T))
MXN.CAD.upperThr <- MXN.CAD.mean.train + 1.96 * MXN.CAD.sd.train
MXN.CAD.lowerThr <- MXN.CAD.mean.train - 1.96 * MXN.CAD.sd.train

# Visualize the spread (percent change) and thresholds
plot(MXN.CAD.spread.train, main = "Mexican Peso vs Canadian Dollar spread (per change) training set", type = "l")
abline(h=MXN.CAD.mean.train, col = "red", lwd = 2)
abline(h=MXN.CAD.upperThr, col="blue", lwd=2)
abline(h=MXN.CAD.lowerThr, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high <- MXN.CAD.spread.train[MXN.CAD.spread.train > MXN.CAD.upperThr]
low <- MXN.CAD.spread.train[MXN.CAD.spread.train < MXN.CAD.lowerThr]
MXN.CAD.buys <- which(MXN.CAD.spread.train %in% high)
MXN.CAD.sells <- which(MXN.CAD.spread.train %in% low)

# Create the spread in price terms
MXN.CAD.spread.price.train <- CurrClose.Train$MXN - MXN.CAD.hr * CurrClose.Train$CAD
# Plot the spread(price) chart with buys and sells marked
plot(MXN.CAD.spread.price.train, main = "Mexican Peso vs Canadian Dollar spread (price) training set", type = "l")
abline(v=MXN.CAD.sells, col='red')
abline(v=MXN.CAD.buys, col='green')
```

#Test the model on Pound v Euro

```{r}
# Test on Pound v Euro 
GBP.EUR.spread.test <- CurrDataPer.Test$GBP - (GBP.EUR.hr * CurrDataPer.Test$EUR)

# compute the statistics of the spread (percent change data)
GBP.EUR.mean.test <- as.numeric(mean(GBP.EUR.spread.test, na.rm=T))
GBP.EUR.sd.test <- as.numeric(sd(GBP.EUR.spread.test, na.rm = T))
GBP.EUR.upperThr.test <- GBP.EUR.mean.test + 1.96 * GBP.EUR.sd.test
GBP.EUR.lowerThr.test <- GBP.EUR.mean.test - 1.96 * GBP.EUR.sd.test

# Visualize the spread (percent change) and thresholds
plot(GBP.EUR.spread.test, main = "Pound vs Euro spread (per change) testing set", type = "l")
abline(h=GBP.EUR.mean.test, col = "red", lwd = 2)
abline(h=GBP.EUR.upperThr.test, col="blue", lwd=2)
abline(h=GBP.EUR.lowerThr.test, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high.test <- GBP.EUR.spread.test[GBP.EUR.spread.test > GBP.EUR.upperThr]
low.test <- GBP.EUR.spread.test[GBP.EUR.spread.test < GBP.EUR.lowerThr]
GBP.EUR.buys.test <- which(GBP.EUR.spread.test %in% high.test)
GBP.EUR.sells.test <- which(GBP.EUR.spread.test %in% low.test)

# Create the spread in price terms
GBP.EUR.spread.price.test <- CurrClose.Test$GBP - GBP.EUR.hr * CurrClose.Test$EUR

# Plot the spread(price) chart with buys and sells marked
plot(GBP.EUR.spread.price.test, main = "Pound v Euro spread (price) testing set", type = "l")
abline(v=GBP.EUR.sells.test, col='red')
abline(v=GBP.EUR.buys.test, col='green')
```

#Test the model on Australian Dollar vs New Zealand Dollar

```{r}

# Create the spread (percent change) on Testing dataset 
AUD.NZD.spread.test <- CurrDataPer.Test$AUD - (AUD.NZD.hr * CurrDataPer.Test$NZD)

# compute the statistics of the spread (percent change data)
AUD.NZD.mean.test <- as.numeric(mean(AUD.NZD.spread.test, na.rm=T))
AUD.NZD.sd.test <- as.numeric(sd(AUD.NZD.spread.test, na.rm = T))
AUD.NZD.upperThr.test <- AUD.NZD.mean.test + 1.96 * AUD.NZD.sd.test
AUD.NZD.lowerThr.test <- AUD.NZD.mean.test - 1.96 * AUD.NZD.sd.test

# Visualize the spread (percent change) and thresholds
plot(AUD.NZD.spread.test, main = "Australia v New Zealand spread (per change) testing set", type = "l")
abline(h=AUD.NZD.mean.test, col = "red", lwd = 2)
abline(h=AUD.NZD.upperThr.test, col="blue", lwd=2)
abline(h=AUD.NZD.lowerThr.test, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high.test <- AUD.NZD.spread.test[AUD.NZD.spread.test > AUD.NZD.upperThr]
low.test <- AUD.NZD.spread.test[AUD.NZD.spread.test < AUD.NZD.lowerThr]
AUD.NZD.buys.test <- which(AUD.NZD.spread.test %in% high.test)
AUD.NZD.sells.test <- which(AUD.NZD.spread.test %in% low.test)

# Create the spread in price terms
AUD.NZD.spread.price.test <- CurrClose.Test$AUD - AUD.NZD.hr * CurrClose.Test$NZD

# Plot the spread(price) chart with buys and sells marked
plot(AUD.NZD.spread.price.test, main = "Australia v New Zealand spread (price) testing set", type = "l")
abline(v=AUD.NZD.sells.test, col='red')
abline(v=AUD.NZD.buys.test, col='green')
```

#Test the model on Chinese Renminbi vs Hong Kong Dollar

```{r}
# Create the spread (percent change) on Testing dataset 
CNY.HKD.spread.test <- CurrDataPer.Test$YUAN - (CNY.HKD.hr * CurrDataPer.Test$HKD)

# compute the statistics of the spread (percent change data)
CNY.HKD.mean.test <- as.numeric(mean(CNY.HKD.spread.test, na.rm=T))
CNY.HKD.sd.test <- as.numeric(sd(CNY.HKD.spread.test, na.rm = T))
CNY.HKD.upperThr.test <- CNY.HKD.mean.test + 1.96 * CNY.HKD.sd.test
CNY.HKD.lowerThr.test <- CNY.HKD.mean.test - 1.96 * CNY.HKD.sd.test

# Visualize the spread (percent change) and thresholds
plot(CNY.HKD.spread.test, main = "Chinese Renminbi vs Hong Kong Dollar spread (per change) testing set", type = "l")
abline(h=CNY.HKD.mean.test, col = "red", lwd = 2)
abline(h=CNY.HKD.upperThr.test, col="blue", lwd=2)
abline(h=CNY.HKD.lowerThr.test, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high.test <- CNY.HKD.spread.test[CNY.HKD.spread.test > CNY.HKD.upperThr]
low.test <- CNY.HKD.spread.test[CNY.HKD.spread.test < CNY.HKD.lowerThr]
CNY.HKD.buys.test <- which(CNY.HKD.spread.test %in% high.test)
CNY.HKD.sells.test <- which(CNY.HKD.spread.test %in% low.test)

# Create the spread in price terms
CNY.HKD.spread.price.test <- CurrClose.Test$YUAN - CNY.HKD.hr * CurrClose.Test$HKD

# Plot the spread(price) chart with buys and sells marked
plot(CNY.HKD.spread.price.test, main = "Chinese Renminbi vs Hong Kong Dollar spread (price) testing set", type = "l")
abline(v=CNY.HKD.sells.test, col='red')
abline(v=CNY.HKD.buys.test, col='green')
```

#Test the model on Mexican Peso v Canadian Dollar

```{r}
# Create the spread (percent change) on Testing dataset 
MXN.CAD.spread.test <- CurrDataPer.Test$MXN - (MXN.CAD.hr * CurrDataPer.Test$CAD)

# compute the statistics of the spread (percent change data)
MXN.CAD.mean.test <- as.numeric(mean(MXN.CAD.spread.test, na.rm=T))
MXN.CAD.sd.test <- as.numeric(sd(MXN.CAD.spread.test, na.rm = T))
MXN.CAD.upperThr.test <- MXN.CAD.mean.test + 1.96 * MXN.CAD.sd.test
MXN.CAD.lowerThr.test <- MXN.CAD.mean.test - 1.96 * MXN.CAD.sd.test

# Visualize the spread (percent change) and thresholds
plot(MXN.CAD.spread.test, main = "Mexican Peso v Canadian Dollar spread (per change) testing set", type = "l")
abline(h=MXN.CAD.mean.test, col = "red", lwd = 2)
abline(h=MXN.CAD.upperThr.test, col="blue", lwd=2)
abline(h=MXN.CAD.lowerThr.test, col="blue", lwd=2)


# Pull the dates of the highest and lowest levels of spread (percent change)
high.test <- MXN.CAD.spread.test[MXN.CAD.spread.test > MXN.CAD.upperThr]
low.test <- MXN.CAD.spread.test[MXN.CAD.spread.test < MXN.CAD.lowerThr]
MXN.CAD.buys.test <- which(MXN.CAD.spread.test %in% high.test)
MXN.CAD.sells.test <- which(MXN.CAD.spread.test %in% low.test)

# Create the spread in price terms
MXN.CAD.spread.price.test <- CurrClose.Test$MXN - MXN.CAD.hr * CurrClose.Test$CAD

# Plot the spread(price) chart with buys and sells marked
plot(MXN.CAD.spread.price.test, main = "Mexican Peso v Canadian Dollar spread (price) testing set", type = "l")
abline(v=MXN.CAD.sells.test, col='red')
abline(v=MXN.CAD.buys.test, col='green')
```

#Summary

We have used the unsupervised method of Hierarchical Clustering to separate currency data into similar pairs. From there we used linear regression modeling to find relationships between the pairs. Although the R-squareds do not give us a tight correlation, we can develop trading rules to take advantage of the dissimilarity in the pairs. When the daily percent change spread deviates outside of the 95% confidence interval of the mean of the spread, we can see trading opportunities on the chart of the price spread.