---
title: "ML Workshop1 Week 5"
author: "John Navarro"
date: "July 21, 2017"
output: pdf_document
---
# Workshop 1 Recursive Binary Splitting

## 1 Example on binary splitting
```{r}
# read data
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/MachineLearning/Week5_Trees"
(tree.data<-read.csv(file=paste(datapath,"binary_splitting_data.csv",sep="/")))
```
### 1.1 preparing functions for sum of squares
Create function
```{r}
oneSplit <- function(myData, myVariable, splitLevel){
  idx<-myData[,myVariable+1]<mySplit
  set1 <- myData[idx==TRUE,]
  set2 <- myData[idx==FALSE]
  y.mean1 <- mean(set1[,1])
  y.mean2 <- mean(set2[,1])
  SS.1 <- sum((y.mean1-set1[,1])^2)
  SS.2 <- sum((y.mean2-set1[,1])^2)
  SSE=SS.1+SS.2
  return(SSE)
  
}

selectSplit <- function(myData){
  x1 <-sort(myData[,2])
  x2 <- sort(myData[,3])
  nData <- length(myData[,1])-1
  splits <- data.frame(x1=median(x1), x2=median(x2))
  splits1.ss <- oneSplit(myData, myVariable, splits$x1)
  splits2.ss <- oneSplit(myData, myVariable, splits$x2)
  SS <- data.frame(splits1.ss, splits2.ss)
  return(SS, splits)
}


x1 <-sort(tree.data[,2])
x2 <- sort(tree.data[,3])
 nData<-length( tree.data[,1])-1
splits<-data.frame(x1=sapply(1:nData,function(z) mean(c(x1[z],x1[z+1]))),
                     x2=sapply(1:nData,function(z) mean(c(x2[z],x2[z+1]))))
```
```{r}
oneSplit<-function(myData,myVariable,mySplit){
  # myData: matrix of data
  # myVariable: number of variable that needs split 1:ncol(myData)
  # mySplit: split level
  idx<-myData[,myVariable+1]<mySplit # separate variable by mySplit 
  subs1<-subset(myData,idx) # subset below mySplit
  subs2<-subset(myData,!idx) # subset above mySplit
  mean1<-mean(subs1$y) # mean of subset 1
  mean2<-mean(subs2$y) # mean of subset 2
  sum((subs1$y-mean1)^2)+sum((subs2$y-mean2)^2) # out: sum of sums of squares
}
selectSplit<-function(myData){
  # myData: data matrix including response in column 1
  x1<-sort(myData[,2]) # sort variable 1
  x2<-sort(myData[,3]) # sort variable 2
  nData<-length(myData[,1])-1 # number of splits
  
  # create splits
  splits<-data.frame(x1=sapply(1:nData,function(z) mean(c(x1[z],x1[z+1]))),
                     x2=sapply(1:nData,function(z) mean(c(x2[z],x2[z+1]))))
  
  # calculate sums of squares of all splits
  splits1.SS<-sapply(1:nData,function(z) oneSplit(myData,1,splits$x1[z]))
  splits2.SS<-sapply(1:nData,function(z) oneSplit(myData,2,splits$x2[z]))
  
  # out: sums of squares and splits
  list(SS=data.frame(splits1.SS=splits1.SS,splits2.SS=splits2.SS),
       splits=splits)
}
```



## 1.2 Find node 1, fist split variable and split level
```{r}
# splits for node 1
(node1<-selectSplit(tree.data))

# Determine which variable needs split? compare minimum SS for each variable
(splitVar<-ifelse(min(node1[["SS"]][,"splits1.SS"])>min(node1[["SS"]]["splits2.SS"]),2,1)) # if 2nd variable is smaller returns 2, if 1st variable is smaller, it returns 1

# where to split? row number corresponding to smallest SS
(splitNum<-as.numeric(ifelse(splitVar==1,which.min(node1[["SS"]][,"splits1.SS"]), which.min(node1[["SS"]][,"splits2.SS"]))))  # if flitVAr is 1, return min of splits1.SS, otherwise return the min of splits2.SS

# What is the split level? row splitNum in column splitVar
(splitLevel.1<-node1[["splits"]][splitNum,splitVar]) # takes which row and which variable, extracts from splits data frame

```


## 1.3 Create 2 regions R1 and R2 breaking the initial sample

```{r}
# What are the regions created? create subsets based on the split
(Reg1<-subset(tree.data,tree.data[,splitVar+1]>=splitLevel.1)) # splitVar=1=x1
(Reg2<-subset(tree.data,tree.data[,splitVar+1]< splitLevel.1))

# Plot the two regions
plot(sort(tree.data$x1), sort(tree.data$x2), type="n")
if (splitVar==1) {
  abline(v=splitLevel.1)
} else{
  abline(h=splitLevel.1)
}
legend("topright", legend=c("Split 1"), lty=1)
```

## Which of the regions R1, R2 needs a split?

```{r}
(node2 <- selectSplit(Reg1)) # This looks like a leaf

# Check if Region 2 needs to be split
(node2 <- selectSplit(Reg2))
# Yes this needs further split
```

## 1.5 Split R2 by creating regions R3 and r4

```{r}
# Determine which variable needs split? compare minimum SS for each variable
(splitVar<-ifelse(min(node2[["SS"]][,"splits1.SS"])>min(node2[["SS"]]["splits2.SS"]),2,1)) # if 2nd variable is smaller returns 2, if 1st variable is smaller, it returns 1
```
## 1.5 What are the split variables and split levels?
```{r}
# where to split? row number corresponding to smallest SS
(splitNum<-as.numeric(ifelse(splitVar==1,which.min(node2[["SS"]][,"splits1.SS"]), which.min(node2[["SS"]][,"splits2.SS"]))))  # if splitVAr is 1, return min of splits1.SS, otherwise return the min of splits2.SS

# What is the split level? row splitNum in column splitVar
(splitLevel.2<-node2[["splits"]][splitNum,splitVar]) # takes which row and which variable, extracts from splits data frame
```

## 1.7 Create the next breakup

```{r}
# What are the regions created? create subsets based on the split
(Reg3<-subset(Reg2,Reg2[,splitVar+1]>=splitLevel.2)) # splitVar=1=x1
(Reg4<-subset(Reg2,Reg2[,splitVar+1]< splitLevel.2))

#plot
plot(sort(tree.data$x1), sort(tree.data$x2), type="n")
if (splitVar==1) {
  abline(v=splitLevel.1)
  abline(v=splitLevel.2,lty=2)
} else{
  abline(h=splitLevel.1)
  abline(h=splitLevel.2,lty=2)
}


legend("topright", legend=c("Split 1", "Split 2"), lty=c(1,2))
```

## 1.8 Which region needs a split
```{r}
# Check if Region 4 needs to be split
(node3 <- selectSplit(Reg4)) # This looks like a leaf

# Check if Region 3 needs to be split
(node3 <- selectSplit(Reg3)) # This needs to be split

```

## 1.9 Select next split variable for R3


```{r}
# Determine which variable needs split? compare minimum SS for each variable
(splitVar<-ifelse(min(node3[["SS"]][,"splits1.SS"])>min(node3[["SS"]]["splits2.SS"]),2,1)) # if 2nd variable is smaller returns 2, if 1st variable is smaller, it returns 1

# where to split? row number corresponding to smallest SS
(splitNum<-as.numeric(ifelse(splitVar==1,which.min(node3[["SS"]][,"splits1.SS"]), which.min(node3[["SS"]][,"splits2.SS"]))))  # if splitVAr is 1, return min of splits1.SS, otherwise return the min of splits2.SS

# What is the split level? row splitNum in column splitVar
(splitLevel.3<-node3[["splits"]][splitNum,splitVar]) # takes which row and which variable, extracts from splits data frame

```

# Create Regions R5 and R6

```{r}
(Reg5 <- subset(Reg3, Reg3[, splitVar+1]>=splitLevel.3))
(Reg6 <- subset(Reg3, Reg3[, splitVar+1]<splitLevel.3))

# Plot the splits
plot(sort(tree.data$x1), sort(tree.data$x2), type="n")
if (splitVar==1) {
  abline(v=splitLevel.1)
  abline(v=splitLevel.2,lty=2)
  abline(v=splitLevel.3,lty=3)
} else{
  abline(h=splitLevel.1)
  abline(h=splitLevel.2,lty=2)
  abline(v=splitLevel.3,lty=3)
}
legend("topright", legend=c("Split 1", "Split 2", "Split 3"), lty=c(1,2,3))

```

## 1.11 Find mean values of all regions
```{r}
#Since all regions except 5  have one observarion, only have to find mean of reg 5
mean(Reg5$y)
(nodeMeans <- sapply(1:6, function(z) mean(get(paste0("Reg",z))$y)))
```

## 1.12 Summary

Initial split is at X1 at level 1.37464
Second split is at X1 at level -0.97632
Third split is at X1 at level -0.3135914

## 1.13 Check the splits against fitted Tree
```{r}
library(rpart)
library(rpart.plot)
built.tree<-rpart(y~., data=tree.data,
                        control = rpart.control(minbucket=1,cp=.0001))
printcp(built.tree)

# return best CP
(best.CP <- built.tree$cptable[which.min(built.tree$cptable[,"xerror"]),"CP"])

# Plot tree
prp(built.tree,extra=101, # display prob of classes in the node
    branch=.5, # change angle of branch lines
    shadow.col="gray", # shadows under the leaves
    branch.lty=3, # draw branches using dotted lines
    split.cex=1.2, # make the split text larger than the node text
    split.prefix="is ", # put "is " before split text
    split.suffix="?", # put "?" after split text
    split.box.col="lightgray", # lightgray split boxes (default is white)
    split.border.col="darkgray", # darkgray border on split boxes
    split.round=.5,
    nn=TRUE)

# Find node mean
nodeMeans
```

