---
title: "Workshop1"
author: "John Navarro"
date: "December 9, 2016"
output: pdf_document
---

1.1 one categorical predictor

```{r}
taste.model<-lm(score~scr,data=tasteData)
summary(taste.model)
```

hypth b1=0, rejected


1.2 2 factors
one way anova with 4 treatments, diff combinations of screen

```{r}
taste.model.4.treatments<-lm(score~type,data=tasteData)
anova(taste.model.4.treatments)
```
type is a categorical variable, its the predictor, it is significant
So not all groups are equal, from anova
```{r}
taste.cint<-confint(glht(taste.model.4.treatments,mcp(type="Tukey")))
taste.cint
```
tukey test, multiple comparison
pairwise comparisons
plot to show which contrasts not equla to zero, DA and DB

1.3 2 factors and interaction

```{r}
taste.lm<-lm(score~scr*liq,data=tasteData)
anova(taste.lm)
```
screen is sign, not liq nor scr:liq
first look at interactions, bc if significant, is there reason to look at main effects? if interatction is significant, then you want to simplyfy
summary(taste.lm)

scrfine slope is significant, not the others

Hypothesis, is B3 =0?

testing contrast one, testing B1+B3/2

go back and look at, p value for interaction in anova
3rd contrast p value is same in summary and anova, that B3 = 0

first 2 are testing diff hypoth
anova testing B1, B2 =0

2 way anova, is testing C1=0,

create linear model with contrast matrix below, replaces default contrast matrix, with one below

```{r}
M<-cbind("C1"=rep(c(-1,-1,1,1),each=4)/2,
         "C2"=rep(c(-1,1,-1,1),each=4)/2,
         "C3"=rep(c(1,-1,-1,1),each=4)/2)
taste.lm2<-lm(score~M,data=tasteData)
summary(taste.lm2)
```
It gives contrast testing
P values are the same as 2 way anova


#Taguchi experiment

3 factors, each 2 levels
(designTaguchi<-oa.design(L4.2.3))

website testing
2 diff factors, new font and new button color, nuisance variable is gender, conversion results day 1

First factorial design
(webSiteTesting.Taguchi<-webSiteTesting.factorial[c(1,3,2,4),])



taguchi

webSiteTesting.lm.Taguchi<-lm(Conversion.day1~Font.Type+Button.Color+Gender,data=webSiteTesting.Taguchi)
summary(webSiteTesting.lm.Taguchi)

has NAs, why?
too many parameters, too few observations, overfitted

anova shows sum of squares are zero, bad

eliminate font type, run anova, font type not significant, which is different from in factorial design.

drop gender, 

add day 2, double taguchi plan

font sshows significance, anova shows, font sig, gender on borderline

fit lm with interactions,

#Multiplex Arrays

10 classes, each is a channel, plot them
create multiplex assay design, need 4 binary decimals, code channesl 1-10
comb 0000, doesnt work bc it has no signal
assumption is one may show signal, but maybe none

use multiplexer to mix them together, response and class for anova

reduce # of predictors, dont want 10. bc decreases acccuracy

anova(multiplexData.lm)$"Pr(>F)"[1]
omnibus hypoth cannt be rejected


mix 2 and mix 4 show contamination, 
channel 5 has the signal, monitoring only 4 channels, fitting 4 predictors


