---
title: "FA_Workshop Week 5"
author: "John Navarro"
date: "July 20, 2017"
output: pdf_document
---

```{r}
library(forecast)
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week5_Cointegration"
x<-read.csv(file=paste(datapath,"WGS1_3YR.csv",sep="/"))
head(x)
dim(x)#1557 x3
r1 <- x[,2]
r3 <- x[,3]
plot(r1,col ="blue",type="l",main = "1 and 3 year constant maturity rates")
lines(r3,col="red",main = "1 and 3 year constant maturity rates")
```



r3t=a+Br1t+Et




## Title:
##  ARIMA Modelling
##
## Call:
##  armaFit(formula = ~arma(0, 1), data = cresiduals)
##
## Model:
##  ARIMA(0,0,1) with method: CSS-ML
##
## Coefficient(s):
##        ma1   intercept
##  0.1576619  -0.0001134
##
## Residuals:
##       Min        1Q    Median        3Q       Max
## -0.535503 -0.032129 -0.003149  0.031209  0.293120 # even distribution
##
## Moments:
## Skewness Kurtosis
## -0.01862  6.74808 Fat tails, not gaussian
##
## Coefficient(s):
##             Estimate  Std. Error  t value Pr(>|t|)
## ma1        0.1576619   0.0238446    6.612 3.79e-11 *** coeff is invertable
## intercept -0.0001134   0.0016731   -0.068    0.946
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
##
## sigma^2 estimated as: 0.00325
## log likelihood:       2249.4
## AIC Criterion:        -4492.79

#standardized residuals, variance doesnt look stationary

qqplot shows fat tails

box ljung says not more than 4 lags have autocorrelation

## Cointegration

### Interest Rates Example for the 900 Last Observations
```{r}
n <- length(r1) #1557
nb <- max(n-900,1) #657
rr1 <- r1[nb:n]
rr3 <- r3[nb:n]
plot(rr1, col="blue", type="l", main="1 and 3 year constant maturity rates")
lines(rr3, col="red")
```
Fit cointegration model
```{r}
library(urca)
data <- cbind(rr1, rr3) 
head(data)
cajo <- ca.jo(data,ecdet="none" ,type="eigen", K=2, spec="longrun")
summary(cajo)

```
Residuals and their ACFs and PACFs for 1 year and 3 year rate respectively
```{r}
plotres(cajo)
```
Check statistics and critical values of the test for cointegration order
```{r}
cajo@teststat
cajo@cval
```

Plot test statistic and critical values
```{r}
barplot(cajo@cval[1,],main = "Johansen test h<=1",col = "red")
abline(h=cajo@teststat[1], col="blue")
legend("topleft", c("critical values","test statistics"), lwd=2,col = c("red","blue"), bty="n")
```
Here we cannot reject the NH, so we accept that the cointegration order is less than or equal to 1.
Now we check the second NH. Here we can reject at the 95% significance level and can say that the order is not equal to zero. 

Cointegration vector a1=(a1,1,a1,2), normalized with respect to the first variable
```{r}
# cajo@V normalizes the vector, select the first column
(a_1<- cajo@V[,1])
# By definitigon of cointegration w order h=1, process must be stationar
z_t1 <- data%*% a_1
matplot(z_t1, type="l", main="z(1,t) = a1'x(t)", col="blue")

```
Estimate autoregression model for process zt,1
```{r}
zar <- ar(z_t1, aic=TRUE, method = "yule-walker")
zar$order
auto.arima(z_t1, max.q=0, max.d=0, max.p=6, max.order=6)
```

Check the roots of characteristic equation
```{r}
library(fArma)
par(mfrow = c(1, 1), cex = 0.9)
armaRoots(zar$ar,lwd = 8, n.plot = 400, digits = 8)
zar$ar
```

Since cointegration order equals 1 vector a2=(a2,1, a2,2) is not a cointegration vector and the process zt2=a2xt is not stationary
```{r}
# normalizes 2nd vector to first element, 
a_2 <- cajo@V[,2]
# data %*% the cointegration vector
z_t2 = data%*%a_2
# plot the cointegrated time series
matplot(z_t2, type="l", main="z(2,t)=a2'x(t)", col = "blue")
```
This is not stationary

## Predicting Using Cointegration Model
Xt can be represented as a 2 dim VAR(2) model
rewrite in matrix form
this is the form in which function ca.jo() fits the modelwith setting spec="longrun"

In order to forecast the fitted model
vector mu:
```{r}
(mu <- cajo@GAMMA[,1])
```
Matrix of coefficients PI 1 (slot PI of class ca.jo)
```{r}
(PI <- cajo@PI)
```
Matrix of coefficients G(slot Gamma of class ca.jo)
```{r}
(Gamma <- cajo@GAMMA[,2:3])
```
In order to construct forecasts we also need matrix delta Xt-1 and matrix Xt-2
```{r}
#first difference of the 1 and 3 yr yields
dX_1 <- cajo@Z0
dim(dX_1)
# slot ZK of ca.jo looks like initial data
X_2 <- cajo@ZK
head(X_2)
head(data)
```
```{r}
deltaX_t_1 <- Gamma%*%t(dX_1) + PI %*%t(X_2)
deltaX_t_1<-apply(deltaX_t_1,2,"+",mu)
```
The forecasts for rates are
```{r}
nrowsdata <- dim(data)[1] #901
data_t_2 <- data[3:nrowsdata,] # dim 899x2
deltaX_t_1 <- t(deltaX_t_1) # dim 899x2
# forecX is original X's plus the change in Xs from the forecast equation
forecX <- data_t_2 +deltaX_t_1 # dim 899 x 2
```

plot
```{r}
fr1 = cbind(rr1[3:length(rr1)],forecX[,1])
matplot(fr1,col =c("blue","red"),type="l",main = "1 year rate and prediction")
legend("topright", c("1 year rate","prediction"), lwd=2,col = c("blue","red"), bty="n")
```
Plot predictions of 3 year yield
```{r}
fr3 = cbind(rr3[3:length(rr3)],forecX[,2])
matplot(fr3,col =c("blue","red"),type="l",main = "3 year rate and prediction")
legend("topright", c("3 year rate","prediction"), lwd=2,col = c("blue","red"), bty="n")
```
Cointegration model preserved long term dependence of 1 and 3 year rates
Difference the forecasts and plot them
```{r}
dfr1=diff(fr1)
dfr3=diff(fr3)
plot(dfr1,dfr3,col ="blue",main = "Scatter plot for change of prediction for 1 and 3 year interest rate",
     xlab="Differenced Forecasts of 1-yr Rate",ylab="Differenced Forecasts of 3-yr Rate")
```
This plot shows that the cointegration model also captured short term dependence of rates differences
Cointegration model in contrast to the regression allows us predicting both 1 yr and 3 yr rate
Check errors of prediction by the cointegration model
```{r}
# Calculate 1 yr yield forecast errors
cerror1 <- rr1[3:length(rr1)] - forecX[,1]
# Calcualte 3 year yield forecast errors
cerror3<-rr3[3:length(rr1)]-forecX[,2]
# Plot 1 yr yield
matplot(cerror1,main = " Error of Prediction of 1-Year Yield",type = "l")
# Plot 3 yr yield
matplot(cerror3,main = " Error of Prediction of 3-Year Yield",type = "l")
# plot scatterplot of the errors
plot(cerror1, cerror3, col="blue", main="Scatter plot for prediction of 1 and 3 year interest rate")
# Estimate the covariance matrix of residuals of the cointegration model
cor(cbind(cerror1, cerror3))

```

## Regression model for the 900 last observations

```{r}
# plot the 900 observations for both yields
plot(rr1, rr3, col="blue", main="Scatter plot of interest rates")

# fit a linear model
linreg_900 <- lm(rr3~rr1)
summary(linreg_900)

# Look at the residuals
residuals_900 <- linreg_900$residuals
plot(residuals_900, type="l", col="blue")
```
The process for residuals is unit root non stationary
```{r}
# ACF plot of the residuals
acf(residuals_900, col="blue", lty=1, lwd=4)
```
This shows us definite autocorrelation. Slow decay. this is non stationary data

What does it tell us about the cointegration vector?
If we normalize the coefficients of the linear model, we get (1,1.169) compared to (1,-1.127) which is our cointegrated vector. We can see that both models are capturing a similar relationship.

The AR nature of residuals makes us consider examing the differenced time series
```{r}
# Calculate and plot the first differences
cc1 <- diff(rr1)
cc3 <- diff(rr3)
plot(cc1,cc3,main="Scatter Plot of Differences", col = "blue")
```
The plot shows that the differences remain highly correlated
```{r}
# Fit a linear model to the differences, with no intercept
clinreg_900 <- lm(cc3~cc1 - 1)
summary(clinreg_900)

# Examine the residuals with plot and ACF
cresiduals_900 <- clinreg_900$residuals
plot(cresiduals_900,type ="l",main = "residuals",col = "blue")
acf(cresiduals_900,main = "ACF of residuals",col ="blue",lty=1 ,lwd = 4)

```
Looking at ACF we identify a MA(1) model for the residuals and modify the lin reg model
```{r}
# Use arma fit to estimate MA(1) model on the residuals
ma1_900 <- armaFit(~ma(1), data=cresiduals_900)
summary(ma1_900, which="all")
```
```{r}
#arma residuals
ma1_900res <- residuals(ma1_900)
# Forecast r3, first need to forecast et
theta1_900 <- ma1_900@fit$coef[1]
a_t_900 <- theta1_900*ma1_900res

# let ft be a forecast for rt,3
x_3_900 <- rr3[-length(rr3)] #900

# forecast = 3 yr rates + slope*(diff 1 yr rate) + a_t from MA1
forec_900 <- x_3_900 + clinreg_900$coefficients*cc1+a_t_900
matplot(cbind(rr3[-1],forec_900),type = "l",col = c("blue","red"),main= "3 Year rate and forecast")
legend("topright", c("3 Year rate","Forecasts"), lwd=2,col = c("blue","red"), bty="n")
```
Check the scatter plot of forecast differences vs the difference 1 year yield
```{r}
# take difference of forecasts
difforec_900 = diff(forec_900)
# bind with differenced 1 yr yield
cr_1<-cbind(difforec_900,cc1[-1])
# create scatter plot of the two
plot (cr_1[,1],cr_1[,2],col = "blue",main = "Differenced Forecast of 3-Year Yield vs. Differenced 1-Year Yields")
```
Compare the errors of the regression model with stationary residuals for 3 yr rate of cointegration model
```{r}
# create the errors from forecast
linregerror_900 <- rr3[-1]-forec_900
# bind 3yr errors from lin reg and cointegration model
errors<-cbind(linregerror_900[-1],cerror3)

#plot the two errors
matplot(errors,type ="l",col = c("orange","blue"),main = "3-Year Yield Errors for Regression and Cointegration Model")
legend("topright", c("regression errors","cointegration errors"), lwd=2,col = c("orange","blue"), bty="n")
```
The variance of the cointegration errors (blue) is lower than the regression errors(orange)
Examine how the errors of the tow models are related to each other
```{r}
plot(errors[,1],errors[,2],col = "blue", 
     main = "Scatter Plot of Regression model Errors vs Cointegration errors",
     xlab="Regression Model Errors", ylab="Cointegration Model Errors")
```
The errors of the two models are negatively correlated, this suggests that mixing the modles together (ensemble) maay reduce the error level
