---
title: "FA Homework Week5"
author: "John Navarro"
date: "July 21, 2017"
output: pdf_document
---
## Analysis of Moody's Bond Yields

This assignment helps understanding stationarity and seasonality of linear models for time series
Consider the monthly yields f Moody's AAA and BAA bonds from exercises 4-6 on page 126. The data are in the file  MYieldsData.csv. Analyze possible types of relationships between the two yield variables using regression model with stationary residuals and cointegration.
What is a valid model for predicting the data?

### 1. Exploratory Analysis

```{r, warning=FALSE, include=FALSE }
# Load dependencies
library(forecast)
library(fArma)
library(urca)
```

```{r}
# Read in the data
datapath <- "C:/Users/JohntheGreat/Documents/MSCA/FinancialAnalytics/Week5_Cointegration"
x<-read.csv(file=paste(datapath,"MYieldsData.csv",sep="/"))

# Print head and dimensions of data
head(x)
dim(x)#1115 x3

# Assign the AAA and BAA constant maturity rates
raaa <- x[,2]
rbaa <- x[,3]

# Plot the two time series
plot(raaa, col="black", type="l", main="AAA and BAA Constant Maturity Rates")
lines(rbaa, col="red")

# Make a scatter plot of raaa vs rbaa to observe the correlation
plot(raaa, rbaa)
```
It appears that these time series are correlated, but when the rates are low, it seems like there is more deviation from the correlation.

### 2. Models

#### 2.1 Regression 

```{r}
# Fit to a linear model and print the summary
linreg <- lm(rbaa~raaa)
summary(linreg)

# Explore the residuals
plot(linreg$residuals)#, type="l")
hist(linreg$residuals)
acf(linreg$residuals)
Box.test(linreg$residuals, lag=12, type="Ljung")
```
The plot and histogram of the residuals from linear regression model do not look like gaussian behavior. The ACF plot has highly significant auto correlation at all lags. The residuals are not stationary. Which is also confirmed by the Ljung-Box test. This is not a good model.

Next, we explore if differencing the data will help

```{r}
# Take the first difference of both time series
daaa <- diff(raaa)
dbaa <- diff(rbaa)

# Create a scatter plot of these differences
plot(daaa, dbaa, main="Scatterplot of yield differences")
```
We still see some correlation, fit another linear regression with no intercept
```{r}
# Fit linear regression model
diff.linreg <- lm(dbaa~daaa-1)
summary(diff.linreg)

# Explore the residuals
plot(diff.linreg$residuals)#, type="l")
hist(diff.linreg$residuals)
acf(diff.linreg$residuals)
Box.test(diff.linreg$residuals, lag=12, type="Ljung")
```
From the residual plot, we see that there are a few outliers while most of the residuals are close to zero. This is confirmed by looking at the histogram. ACF test shows significant correlation at multiple lags. While Ljung-Box confirms that the residuals are not stationary.

Lets try to fit an ARMA model to the residuals

```{r}
dlr.ts <- ts(diff.linreg$residuals, frequency = 12) #1114

# Run residuals through auto.arima function
auto.arima(dlr.ts, seasonal=T)
```
auto.arima suggest an ARMA(2,2) model for the residuals
```{r}
# Fit an ARMA model to the residuals
ma22 <- armaFit(~arma(2,2), data=dlr.ts)
summary(ma22, which="all")
```

############# Does this ARMA fit look good enough?##############

Build forecast for BAA rate using linear model plus arma for erros

##############what equation am I using to forecast?##############
My residuals should be Xt, Xt-1 etc, what do I use for et, et-1 in the ARMA2,2 equation....

```{r}
#Extract residuals
ma22_res <- residuals(ma22) #1114
et2 <- ma22_res
et1 <- ma22_res[-1]
et <-et1[-1]

head(et) #1112
head(et1)#1113
head(et2)#1114

length(et2)


# Create Xt-1and Xt-2
xt2 <- dlr.ts[-length(dlr.ts)]
xt1 <- dlr.ts[-1]
head(xt1) #1113
head(xt2) #1113

# Extract ARMA 2,2 coefficients
phi1 <- ma22@fit$coef[1]
phi2 <- ma22@fit$coef[2]
theta1 <- ma22@fit$coef[3]
theta2 <- ma22@fit$coef[4]
a_t <- phi1*xt1 + phi2*xt2 + et + theta1*et1+theta2*et2
head(a_t)
```
Forecast one step ahead for BAA
```{r}
x_B <- rbaa[-length(raaa)]
#forecast <- BAArates(-1) + slope * differenced AAA + error from arma22
forec <- x_B + diff.linreg$coefficients*daaa +a_t
# plot the BAA yields and the forecast
matplot(cbind(rbaa[-1],forec),type = "l",col = c("blue","red"),main= "3 Year rate and forecast",ylab="3Yr Rate and Forecast")
legend("topright", c("BAA rates","Forecasts"), lwd=2,col = c("blue","red"), bty="n")
```

Check a scatter plot of BAA forecast differences vs AAA differences
```{r}
# Shorten daaa to match length  of difforec
daaa.2 <- daaa[-1]
difforec <- diff(forec) #1113

# Bind the two time series into a matrix
cr <- cbind(difforec, daaa.2)

# Scatter plot of differenced forecasts of BAA and differenced AAA Yields
plot(cr[,1],cr[,2], col = "blue",main = "Differences of Forecasted BAA vs Differences of AAA Yields",
      xlab="Difference of Forecasts of BAA Rate",ylab="Differences of AAA Rate")
```

#### 2.2 Cointegration

Fit cointegration model
```{r}
# merge AAA and BAA time series
data <- cbind(raaa, rbaa)

#use Johansen Procedure
cajo <- ca.jo(data, ecdet = "none", type="eigen", K=2, spec ="longrun")
summary(cajo)

```
Plot residuals and ACFs and PACFs
```{r}
plotres(cajo)


```
Check Statistics and critical values for the test for cointegration order
```{r}
cajo@teststat
cajo@cval
```
Plot test of null hypothesis
```{r}
barplot(cajo@cval[1,],main = "Johansen test h<=1",col = "red")
abline(h=cajo@teststat[1], col="blue")
legend("topleft", c("critical values","test statistics"), lwd=2,col = c("red","blue"), bty="n")
```
Here we cannot reject the null hypothesis that the cointegration order is less than or equal to one.
```{r}
barplot(cajo@cval[2,],main = "Johansen test h=0",col = "red", ylim=range(0:25))
abline(h=cajo@teststat[2], col="blue")
legend("topleft", c("critical values","test statistics"), lwd=2,col = c("red","blue"), bty="n")
```
Here we can reject the null hypothesis that the cointegration order = 0. Merging both tests together, we can conclude that the Time series have a cointegration order of 1.

```{r}
# Cointegration vector
(a_1 <- cajo@V[,1])

# multiply the data by the cointegration vector
z_t1= data %*% a_1
matplot(z_t1,type ="l", main = "z(1,t)=a1'x(t)", col = "blue")
```
Estimate autoregression model for process zt1
```{r}
zar <-ar(z_t1,  aic = TRUE,method = "yule-walker")
zar$order
```
Check the roots of the characteristic equations
```{r}
par(mfrow = c(1, 1), cex = 0.9)
armaRoots(zar$ar,lwd = 8, n.plot = 400, digits = 8)
```
Check cointegration vector 2, 
```{r}
a_2<- cajo@V[,2]
z_t2= data %*% a_2
matplot(z_t2,type ="l", main = "z(2,t)=a2'x(t)", col = "blue")
```

### Predicting using Cointegration model
