---
title: "FA Homework 6"
author: "John Navarro"
date: "July 28, 2017"
output: pdf_document
---

A stock index is currently at 810 and has volatility of 20%.
The risk-free rate is 5% per year.
Assume that dividend yield equals 1%
RQuantLib.
```{r}
# Install packages
library(RQuantLib)
```
## 1. Price European 6-month call option with strike 800 using EuropeanOption() from 

```{r}
#Create and Assign variables
hwUnderlying <- 810
hwVolatility <- .20
hwRate <- .05
hwDivYield <- .01
hwStrike <- 800
hwMaturity <- .5

# Use EuropeanOption to calculate Call value and greeks
hwcall <- EuropeanOption(type="call", underlying = hwUnderlying, strike=hwStrike,
                         dividendYield=hwDivYield, riskFreeRate=hwRate, maturity=hwMaturity,
                         volatility=hwVolatility)

# Return the call's price
hwcall$value # 58.73271
```

## 2. Calculate the same premium manually using the formulas on the last slide of the lecture notes
C=S0N(d1) - (K*exp(-RT))N(d2) ### what is N here? Normal distribution?
d1=(ln(S0/k) + (R + .5sigma^2)*T)/(sigma*sqrt(T))
d2=d1 - (sigma*sqrt(T))

```{r}
d1=(log(hwUnderlying/hwStrike) + (hwRate + .5*(hwVolatility^2))*hwMaturity)/(hwVolatility*sqrt(hwMaturity))

d2 = d1 - (hwVolatility*sqrt(hwMaturity))

C = (hwUnderlying * d1) - ((hwStrike*exp(-hwRate * hwMaturity))*d2)

C #120.3204

############This is not close to hwcall$value 58.73271
```

Think how dividend yield should affect option price.
The higher the dividend yield, the lower the call price. 

###########Why is this#######################

## 3. Calculate the premium of put option with the same expiry and same strike using put-call parity

Put = Call + Strike*exp(-RT) - Underlying
```{r}
# Use hwcall$value # 58.73271 from Part 1

# Use Put Call Parity equation
hwput.parity = hwcall$value + (hwStrike * exp(-hwRate*hwMaturity)) - hwUnderlying

# Return the put's price
hwput.parity # 28.98064

# Use EuropeanOption to check put value
hwput <- EuropeanOption(type="put", underlying = hwUnderlying, strike=hwStrike,
                         dividendYield=hwDivYield, riskFreeRate=hwRate, maturity=hwMaturity,
                         volatility=hwVolatility)

# Return the put's price
hwput$value #33.02053

############### Why are these different?###################3333333

```

